A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-02 tex/codatadp.tex
more

@@

  * codatatypes: from a theoretical and implementational point of view, like
      datatypes but:
    * codatatypes are never empty (e.g. finite streams are rejected)
    * no acyclicity (e.g. xs = lcons(0, xs) is satisfiable)
    * instead: bisimilarity check
    * (also: ``enumeration'', e.g. of streams)

@@

* monomorphic (or many-sorted) first-order logic
* Signature: $\Sigma = (\Types, \Funcs)$

@@

We assume a signature $\Sigma = (\Types, \Funcs)$.

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes. To every
argument corresponds a selector. The introduced names for the (co)datatypes, the
constructors, and the selectors must be distinct and different from already
introduced ones. Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{11}(\bigl[\const{s}_{111}{:}\bigr]\; \tau_{111}, \,\ldots,\, \bigl[\const{s}_{11n_{11}}{:}\bigr]\; \tau_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m1}{:}\bigr]\; \tau_{1m1}, \,\ldots,\, \big[\const{s}_{1mn_{1m}}{:}\big]\; \tau_{1mn_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{ij} : \tau_{ij1}\times\cdots\times\tau_{ijk_{ij}} \to \delta_i$
and $\const{s}_{ijk} : \delta_i \to \tau_{ijk}$. Defaults are assumed for
the selector names if they are omitted.

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall
outside this fragment. For example, the specification
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{forest}) \\[-\jot]
      \keyw{and}\;\, \ty{forest} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree},\, \ty{forest})
\end{aligned}
\]
is supported, whereas
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{llist}(\alpha) & \,=\, \const{LNil} \,\mid\, \const{LCons}(\alpha,\, \ty{llist}(\alpha)) \\[-.5\jot]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{llist}(\ty{tree}))
\end{aligned}
\]
is beyond the fragment considered here. In principle, rank-1 (top-level)
polymorphism would be easy to support; nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case. So the only
genuinely interesting cases missing are mixed nested (co)recursion
(as in the \ty{tree} example above) as well as (co)recursion through a
non-(co)datatype (both of which make sense \cite{blanchette-et-al-2014-codata}).

@@

\rem{Discuss signature $\Sigma$, theory $\thD$}
This section presents a calculus for determining the satisfiability of conjunctions of $\thD$-constraints.

Our calculus consists of derivation rules that operate on a set $\Ec$ of equalities and disequalities between $\Sigma$-terms.
We will write $\tEc$ to denote the set of $\Sigma$-terms occurring in $\Ec$.
We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sigma$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $s^i_j( \const{C}_j( \vec t ) )$
are replaced by $t_i$, and moreover assume that all additional constraints added to $\Ec$ are normalized in this manner.

>>>>

\newcommand\Sig{\mathrm{\Sigma}}

@@

  * codatatypes: from a theoretical and implementational point of view, like
      datatypes but:
    * codatatypes are never empty (e.g. finite streams are rejected)

\[
      \keyw{codatatype}\; \,\ty{llist} \,=\, \const{LNil} \,\mid\, \const{LCons}(\ty{int},\, \ty{llist})
\]

    * no acyclicity (e.g. xs = lcons(0, xs) is satisfiable)
    * instead: bisimilarity check
    * (also: ``enumeration'', e.g. of streams)

@@

* monomorphic (or many-sorted) first-order logic
* Signature: $\Sig = (\Types, \Funcs)$

@@

We assume a signature $\Sig = (\Types, \Funcs)$.

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Each datatype $\delta$ is equipped by
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
plain, be or be one of the newly introduced (co)datatypes. To every
argument corresponds a selector. The introduced names for the (co)datatypes, the
constructors, and the selectors must be distinct and different from already
introduced ones. Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \const{C}_{1}\rlap{${}_{1}$}\phantom{{}_{m}}(\bigl[\const{s}_{111}{:}\bigr]\; \tau_{111}, \,\ldots,\, \bigl[\const{s}_{11n_{11}}{:}\bigr]\; \tau_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\,\vdots} \\[-\jot] % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\,\mid\,}}\; \const{C}_{1m}(\bigl[\const{s}_{1m1}{:}\bigr]\; \tau_{1m1}, \,\ldots,\, \big[\const{s}_{1mn_{1m}}{:}\big]\; \tau_{1mn_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{ij} : \tau_{ij1}\times\cdots\times\tau_{ijk_{ij}} \to \delta_i$
and $\const{s}_{ijk} : \delta_i \to \tau_{ijk}$. Defaults are assumed for
the selector names if they are omitted.

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall
outside this fragment. For example, the specification
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{forest}) \\[-\jot]
      \keyw{and}\;\, \ty{forest} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree},\, \ty{forest})
\end{aligned}
\]
is supported, whereas
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{pllist}(\alpha) & \,=\, \const{PLNil} \,\mid\, \const{PLCons}(\alpha,\, \ty{pllist}(\alpha)) \\[-.5\jot]
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}(\ty{int},\, \ty{pllist}(\ty{tree}))
\end{aligned}
\]
is beyond the fragment considered here. In principle, rank-1 (top-level)
polymorphism would be easy to support; nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case. So the only
genuinely interesting cases missing are mixed nested (co)recursion
(as in the \ty{tree} example above) as well as (co)recursion through a
non-(co)datatype (both of which make sense \cite{blanchette-et-al-2014-codata}).

@@

\rem{Discuss signature $\Sig$, theory $\thD$}
This section presents a calculus for determining the satisfiability of conjunctions of $\thD$-constraints.

Our calculus consists of derivation rules that operate on a set $\Ec$ of equalities and disequalities between $\Sig$-terms.
We will write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $s^i_j( \const{C}_j( \vec t ) )$
are replaced by $t_i$, and moreover assume that all additional constraints added to $\Ec$ are normalized in this manner.
